---
title: "BI Faculty Data Project"
output:
  word_document: default
  html_document: default
date: '2023-05-17'
---

```{r}
library(tidyverse) 
library(MASS)
```


```{r}
library(readxl)
Faculty_Data <- read_excel("/Users/rag/Desktop/BI/project/Faculty_Data.xlsx")
Faculty_Data
```
#Cleaning the data: 

```{r}
is.na(Faculty_Data)
```

```{r}
MISSING <- is.na(Faculty_Data$ID) |
           is.na(Faculty_Data$Rank)  |
           is.na(Faculty_Data$Experience)  |
           is.na(Faculty_Data$Salary) 
           

# Count the number of rows flagged for deletion
sum(MISSING)
```

There are three missing values in the Faculty Dataset which we will clean by elimination. 

```{r}
as.numeric(Faculty_Data$Salary)
```

```{r}
as.numeric(Faculty_Data$Experience)
```
```{r}
as.character(Faculty_Data$Rank)
```

```{r}
Faculty_Data_NA <- na.omit(Faculty_Data)
is.na(Faculty_Data_NA)
```


```{r}
df=data.frame(Faculty_Data_NA)
MISSING <- is.na(df$ID) |
           is.na(df$Rank)  |
           is.na(df$Experience)  |
           is.na(df$Salary) 
           
# Count the number of rows flagged for deletion
sum(MISSING)
```

```{r}
Faculty_Data_NA
```
Now that our dataset does not have any missing values we can proceed with cleaning the data.
We will check if there are any duplicated values:

```{r}
n_occur <- data.frame(table(Faculty_Data_NA$ID))
n_occur[n_occur$Freq > 1,]
```

```{r}
unique(Faculty_Data_NA)
```

We have cleaned the data by only showing the unique values as there were three duplicated values and they were excluded.

Now we will check if there are any outliers:

```{r}





# Create a boxplot of the dataset, outliers are shown as two distinct points
boxplot(Faculty_Data_NA$Experience, main="Experience")$out
boxplot(Faculty_Data_NA$Salary, main="Salary")$out


```

```{r}
boxplot(Faculty_Data_NA$Experience, plot=FALSE)$out
boxplot(Faculty_Data_NA$Salary, plot=FALSE)$out


```

We found three outliers in the Salary column and two outliers in the Experience column, and now we will remove them


```{r}
Q <- quantile(Faculty_Data_NA$Salary, probs=c(.25, .75), na.rm = FALSE)
iqr <- IQR(Faculty_Data_NA$Salary)
up <-  Q[2]+1.5*iqr # Upper Range  
low<- Q[1]-1.5*iqr # Lower Range
```



```{r}
eliminated <- subset(Faculty_Data_NA, Faculty_Data_NA$Salary > (Q[1] - 1.5*iqr) & Faculty_Data_NA$Salary < (Q[2]+1.5*iqr))
```

```{r}
boxplot(eliminated$Salary, plot=FALSE)$out
```


```{r}
boxplot(eliminated$Salary, main="Salary")$out
```



The salary outliers were removed and now we will remove the Experience's

```{r}
Q <- quantile(Faculty_Data_NA$Experience, probs=c(.25, .75), na.rm = FALSE)
iqr <- IQR(Faculty_Data_NA$Experience)
up <-  Q[2]+1.5*iqr # Upper Range  
low<- Q[1]-1.5*iqr # Lower Range
```


```{r}
eliminated<- subset(Faculty_Data_NA, Faculty_Data_NA$Experience > (Q[1] - 1.5*iqr) & Faculty_Data_NA$Experience < (Q[2]+1.5*iqr))
```

```{r}
boxplot(eliminated$Experience, plot=FALSE)$out
```


```{r}
boxplot(eliminated$Experience, main="Experience")$out
```

Some of the Ranks are misspelled and we will correct them by using mutate:


```{r}
eliminated <- mutate(eliminated, Rank = recode(.x=Rank,"AssocProff" = "AssocProf"))
eliminated <- mutate(eliminated, Rank = recode(.x=Rank,"AssstProf" = "AsstProf"))
```

```{r}
qplot(x=Experience, y=Salary, data=eliminated,
      color = Rank,
      shape = Rank,
      xlab = "Experience (years)",
      ylab = "Salary (dollars)") 
```



```{r}
fa <- scale (eliminated$Salary)     
Outlier <- which(abs(fa) > 3) 
eliminated <- eliminated[-Outlier, ] 
```
The dataset is now clean and ready to be plotted. 


##Plotting the Experience:
 
```{r}
qplot(x = Experience, data = eliminated, geom = "bar")

```
 
```{r}

base.plot <- ggplot(eliminated, aes(x = Experience)) +
  xlab("Experience") 
base.plot + geom_histogram()

```


```{r}
base.plot + geom_histogram(aes(fill = Experience))
```

#Plotting the Salary:

```{r}
base.plot <- ggplot(eliminated, aes(x = Salary)) +
  xlab("Salary") 
base.plot + geom_histogram()
```




#Plotting the Rank:

```{r}
base.plot <- ggplot(eliminated, aes(x = Rank)) +
  xlab("Rank") 
base.plot + geom_histogram(stat="count")
```


#Plotting The Whole Data: 

```{r}
par(mfrow = c(2,2)) # Display plots in a single 2 x 2 figure 
plot(eliminated$Experience)
with(eliminated, hist(Experience))
plot(eliminated$Salary)
with(eliminated, hist(Salary))
```


```{r}
with(eliminated, plot(Experience, Salary))
```

```{r}
with(eliminated, plot(Experience, Salary))
```
```{r}
library(ggplot2)
```


```{r}
qplot(Experience, Salary, data=eliminated)  # using qplot from ggplot2
```



```{r}
qplot(Rank, Salary, data=eliminated)  # using qplot from ggplot2
```


```{r}
ggplot(eliminated, 
       aes(x=Salary, y=Experience, shape=Rank, color=Rank)) + 
  geom_point() + # Adds points (scatterplot)
  geom_smooth(method = "lm") + # Adds regression lines
  ylab("Salary (dollars)") + # Changes y-axis label
  xlab("Experience (years)") + # Changes x-axis label
  ggtitle("Salary by Experience") # Changes plot title
```

```{r}
library(ggplot2)
library(gridExtra)
 
# Make 3 simple graphics:
g1 <- ggplot(eliminated, aes(x=Salary)) + geom_density(fill="slateblue")
g2 <- ggplot(eliminated, aes(x=Experience, y=Salary, color=Rank)) + geom_point(size=5) + theme(legend.position="none")
g3 <- ggplot(eliminated, aes(x=factor(Rank), y=Salary, fill=Rank)) + geom_boxplot() + theme(legend.position="none")
g4 <- ggplot(eliminated , aes(x=factor(Rank), fill=factor(Rank))) +  geom_bar()
 
# Plots
grid.arrange(g2, arrangeGrob(g3, g4, ncol=2), nrow = 2)
grid.arrange(g1, g2, g3, nrow = 3)
grid.arrange(g2, arrangeGrob(g3, g4, ncol=2), nrow = 1)
grid.arrange(g2, arrangeGrob(g3, g4, nrow=2), nrow = 1)
```

